---
title: "Gender Pay Gap - Review of Data Quality"
output: html_notebook
---

```{r set_up}
library(tidyverse)
library(here)
library(naniar)
library(finalfit)

# import clean data
paygap <- read_rds(here('gender_paygap/data/processed/paygap_clean.rda'))
```

# Timeliness

Most public authority employers must use a snapshot date of 31 March. They must report and publish their gender pay gap information by 30 March of the following year

Private, voluntary, and all other public authority employers must use a snapshot date of 5 April. They must report and publish their gender pay gap information by 4 April of the following year.

It is a legal requirement for all employers who are required to report and publish their gender pay gap information to accurately report this information. Failing to do this within one year of the snapshot date is unlawful. The Equality and Human Rights Commission has the power to enforce any failure to comply with the regulations. Employers that fail to report on time, or report inaccurate data, will be in breach of the regulations and risk facing enforcement action from the Equality and Human Rights Commission (EHRC), leading to court orders and fines.

```{r timeliness}
paygap %>% 
    summarise(min = min(delay),
              mean = mean(delay),
              median = median(delay),
              max = max(delay)) %>% 
    round(1) 

paygap %>% 
    ggplot(aes(delay)) +
    geom_histogram(bins = 30, ) +
    geom_vline(xintercept = - 366, 
               colour = 'red',
               linetype = 'dashed')

paygap %>% 
    ggplot(aes(delay)) +
    geom_boxplot(outlier.alpha = 0.1) +
    geom_vline(xintercept = - 366, 
               colour = 'red',
               linetype = 'dashed')

sum(paygap$late_submission)
round(mean(paygap$late_submission) * 100)

length(unique(paygap[paygap$delay < -366, ]$employer_id))
```

Most employers appear submit their data just before their deadline. However 0n 6764 (\~14%) occasions data was submitted late. The data shows that there is evidence of some employers submitting data over 1500 days late. As well as being unlawful long delays could have implications for the accuracy of data.

There were 96 occasions where employers appear to have submitted data over a year before their due date - therefore before their snapshot date. This calls into question the validity of this data.

# Uniqueness

```{r uniqueness}

# check for duplication of observations
sum(duplicated(paygap))

# is employer_id a unique identifier
nrow(paygap) / n_distinct(paygap$employer_id)
# no there are nearly 4 times as many observations as employer ids

# visualise 
paygap %>% 
  count(employer_id) %>% 
  ggplot() +
  geom_histogram(aes(n), binwidth = 0.5)

# most employer ids appear to be associated with multiple records 
# is there any evidence of >1 entry by employer in same year
paygap %>% 
  group_by(employer_id, year) %>% 
  count() %>%
    filter(n > 1) %>% 
    nrow()
```

There are no duplicate records in this data set. There are multiple observations per employer id due to data being submitted across multiple years. There is only one observation per employer id:year pair - making this pairing a unique identifier. There were 4 occasions identified in the raw data where employers had entered 2 submissions for the same year. This was dealt with during data cleaning where only the most recent entry was retained.

# Validity

There were a number of areas to explore around validity of the data

### Early submission

```{r early_submit}

# early data submission

paygap %>% 
    filter(delay < -366) %>% 
    group_by(snapshot_date) %>% 
    summarise(
        n = n(),
        min_delay = min(delay),
        mean_delay = mean(delay),
        max_delay = max(delay)
    )
```

It was identified during assessment of timeliness that on 96 occasions employers had submitted gender pay gap data prior to their 'snapshot' date. Further investigation revealed that these submissions were all in 2021 where the due date was 6 months later than usual. This coincides with the covid-19 pandemic and may have been the result of relaxation of due dates for reporting due to this. I feel that these submissions are an accurate reflection of the situation on the snapshot day in 2019 and no action needs to be taken.

### Bonus percent

```{r bonus_percent}
paygap %>%
  select(male_bonus_percent, female_bonus_percent) %>%
 summary()

paygap %>% 
    filter(female_bonus_percent > 100) %>% 
    select(current_name, year, female_bonus_percent)
# one employer reported 100.4% of females received a bonus
```

This refers to the percentage of men and women receiving bonus pay. No values should be less than 0% or greater than 100%. All values were valid except for one employer who reported that 100.4% of women had been paid a bonus in 2021

### Difference in hourly percent

```{r diff_hourly_percent}
paygap %>% 
  select(diff_mean_hourly_percent, diff_median_hourly_percent) %>% 
  summary()

paygap %>% 
  ggplot(aes(y = diff_mean_hourly_percent)) +
  geom_boxplot()

paygap %>% 
  ggplot(aes(y = diff_median_hourly_percent)) +
  geom_boxplot() 
```

Positive values indicate that males paid more than females per hour whilst negative means that females paid more than males.

The minimum value for median pay difference between males and females was around minus 500% - ie. women paid 5 times more per hour than men. The maximum value was around 120%. There are a number of outlying values - maybe calculate proportion values which are outliers?

### Difference in bonus percent

```{r diff_bonus_percent}
paygap %>% 
  select(diff_mean_bonus_percent, diff_median_bonus_percent) %>% 
  summary()

paygap %>% 
  ggplot(aes(y = diff_mean_bonus_percent)) +
  geom_boxplot()

paygap %>% 
  ggplot(aes(y = diff_median_bonus_percent)) +
  geom_boxplot() 
```

Positive values indicate that males paid more than females as a bonus whilst negative means that females paid more than males.

The minimum value for median pay difference between males and females was around minus 120000 %. The maximum value was 4000%.

These figures are extreme - especially where females paid 120000% more. Explore these employers in more detail. Investigation provides no clear evidence regarding validity - all values are possible. In real world situation these extreme values could be double checked with submitting organisation.

## Hourly pay quartiles

```{r hourly_pay_quartiles}
paygap %>% 
  select(contains('quartile')) %>% 
  summary()

# these are percentages and all are valid ie. between 0-100%
# also average male plus average female values = 100% for each quartile
```

All hourly pay quartile data appears to be valid

## Males / females receiving bonus

# Completeness

## Explicit missing data

```{r}
paygap_selected <-  paygap %>%
    select(-c(
        company_link_to_gpg_info,
        company_number,
        responsible_person,
        address
    ))

paygap_selected %>% 
    pct_complete_case()

paygap_selected %>% 
    pct_miss_case()

paygap_selected %>% 
    pct_miss()

paygap_selected %>% 
    gg_miss_var(show_pct = TRUE)

```

When unimportant variables removed from data 73% of rows are complete leaving 27% having some missing data. 2.0% of all data is missing. The most frequent missing data is for difference in bonus percent followed by sic codes, employer size and then hourly pay quartiles.

### Difference in bonus percent

```{r diff_bonus_percent_NA}

sum(is.na(paygap$diff_median_bonus_percent))
sum(is.na(paygap$diff_mean_bonus_percent))
sum(is.na(paygap$male_bonus_percent))
sum(is.na(paygap$female_bonus_percent))

## employers who don't pay a bonus to one or both sexes
paygap %>% 
  filter(female_bonus_percent == 0 | male_bonus_percent == 0) %>% 
  summarise(missing_mean = sum(is.na(diff_mean_bonus_percent)),
            missing_median = sum(is.na(diff_median_bonus_percent)))

# missing values where bonuses are paid to both sexes
paygap %>%
  # select observations where bonuses are paid to males and females but there
  # is still missing data for difference in bonus percent
  filter((female_bonus_percent != 0 & male_bonus_percent != 0) &
           (is.na(diff_median_bonus_percent) | is.na(diff_mean_bonus_percent)
            )
         )%>%
  nrow() 
```

The largest amount of missing data was for percent difference in bonus pay between males and females. There were 8981 observations with missing values. There were no missing values for percentage of male or female employees receiving bonus.

It appears that most of missing values (8971) relate to employers where bonuses were not paid to either (or both) males or females and therefore difference cant be calculated. There were 14 observations where bonuses were paid to males and females but the value for percent difference in bonus pay was missing.

### Hourly pay quartiles

```{r}
paygap %>% 
  select(contains('quartile')) %>%
  summary()

paygap %>% 
  select(contains('quartile')) %>%
  # plot missingness by variable and observation
  missing_plot()
```

392 observations have missing data relating to hourly pay quartile data. Missing values across multiple quartiles are in same observations. ie. there is either data for all quartiles or for none

### Employer size

```{r}
paygap_selected %>% 
    select(where(any_na)) %>% 
    gg_miss_var(show_pct = TRUE, facet = employer_size)
```

It appears that as employer size increases the percent of missing data for all variables appears to decrease. The percentage of missing data for all variables appears to be highest where employer size is unknown. This suggests that some employers have not provided complete information. Is there a pattern to this - ie. what type of missing data is this?

This is an area to revisit prior to any modelling that requires management of missing data.

## Implicit missing data

This is missing data where there is no row in the dataset for a particular employer:year pairing.

Companies with size 250-499 may fall out of reporting requirements due to changes in workforce size or may go out of business.

Missing data submission most likely to be an issue where there are gaps between years of submission esp if a large company that is unlikely to have reduced its workforce to below 250 employees.

```{r}
paygap %>% 
    select(employer_id, snapshot_date, due_date, year, employer_size) %>% 
    group_by(employer_id) %>% 
    mutate(gap = year - lag(year)) %>% 
    filter(gap > 1) %>% 
    ungroup() %>% 
    count(snapshot_date, year)

paygap %>% 
    select(employer_id, current_name, snapshot_date, due_date, year, employer_size) %>% 
    group_by(employer_id) %>% 
    mutate(gap = year - lag(year)) %>% 
    filter(gap > 1 & employer_size != '250 to 499') %>% 
    ungroup() %>% 
    arrange(desc(employer_size))
```

Some work to do on interpretation of this data
